---
title: "Paleobuddy overview"
author: Bruno do Rosario Petrucci
output: github_document
vignette: >
  %\VignetteIndexEntry{overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(dpi=300)
```

`paleobuddy` is an R package for species birth-death simulation, complete with the possibility of generating phylogenetic trees and fossil records from the results. The package offers unprecedented flexibility in the choice of speciation, extinction, and fossil sampling rates, as we will showcase in this vignette. 

One of the biggest reason to write and publish a simulator is to effectively test rate estimation methods with scenarios whose true dynamics are known---others might include model adequacy tests and the study of scenarios without an analytical solution. This leads to an intuitive overview of the package: in this vignette, we will first generate a couple of useful scenarios to show use cases of the package. Then, we will discuss the choice of rates further, detailing the customization capabilities of both the birth-death and sampling functions. Finally, we will conclude going over the shortcomings of the package, including the features we plan to implement in the future. As new versions are released, this document will be updated to reflect the newest features.

First we do some setup.

```{r results = "hide", message = FALSE}
# importing the package functions
library(paleobuddy)
```

## Constant rate birth-death 

Let us try the simplest possible birth-death scenario - constant speciation and extinction rates.

`bd.sim` is the birth-death simulator in the package. We use it to generate a group.

```{r}
# we set a seed so the results are reproducible
set.seed(1)

# set the necessary parameters
# initial number of species
n0 <- 1

# speciation rate - approx. 1 speciation event every 4my
# we are trying to create a big phylogeny so phytools can function better
lambda <- 0.25

# extinction rate - approx. 1 extinction event every 10my
mu <- 0.15

# maximum simulation time - species that die after this are considered extant
tMax <- 50

# run the simulation
sim <- bd.sim(n0, lambda, mu, tMax = tMax)

# take a look at the way the result is organized
sim
```

The output of `bd.sim` is a `sim` object, a class made up of named vectors that is organized as follows

  * `TE` a vector of extinction times. For an extant species, the extinction time is `NA`.
  * `TS` a vector of speciation times. For species alive at the beginning of the simulation, the speciation time is `tMax`.
  * `PAR` a vector of parents. The naming of species follows the order of `TE` and `TS`, i.e. if `PAR[i] == j`, the species whose speciation time is `TS[j]` generated species `i`. For species alive at the beginning of the simulation, the parent is `NA`.
  * `EXTANT` a logical vector indicating whether a species is alive at the end of the simulation or not. Note this could be extrapolated from the information in `TE`, but we present it for practicality's sake.
  
We can use the function `draw.sim` to visualize the longevity of species in this simulation.

```{r}
# draw simulation
draw.sim(sim, showLabel = FALSE)
```

Species are drawn in order of speciation time by default, though that can be altered. We omit species labels since for a high number of species that can get unruly.

Using this sim object, we can generate a phylogenetic tree using `make.phylo`.

```{r}
# there are currently not many customization options for phylogenies
phy <- make.phylo(sim)

# plot it with APE - hide tip labels since there are a lot so it looks cluttered
ape::plot.phylo(phy, show.tip.label = FALSE)
ape::axisPhylo()

# plot the molecular phylogeny
ape::plot.phylo(ape::drop.fossil(phy), show.tip.label = FALSE)
ape::axisPhylo()
```

From here, we could run this phylogeny through a number of inference software in the field, so as to test their accuracy and robustness. Of course this would require more trees, and larger trees, to control for stochasticity.

For illustration, we create a simulation with more than 500 species, with 253 extant species.

```{r}
# set a seed
set.seed(3)

# create simulation
# note nExtant, defining we want 200 or more extant species at the end
sim <- bd.sim(n0, lambda, mu, tMax = tMax, nExtant = c(200, Inf))

# check the number of extant species
paste0("Number of species alive at the end of the simulation: ", 
       sum(sim$EXTANT))
```

And we could of course get a molecular phylogeny from it.

```{r}
# might look a bit cluttered
ape::plot.phylo(ape::drop.fossil(make.phylo(sim)), show.tip.label = FALSE)
ape::axisPhylo()
```

## Time-dependent speciation and extinction

One of the pluses of `paleobuddy` is that we can generate both fossil records and phylogenies in independent processes, both coming from the same underlying birth-death simulations. We will here use the fossil record generating functions of `paleobuddy` to generate a fossil record and prepare the output for use with PyRate (Silvestro et al 2014) and Foote's Per Capita method (Foote 2000), as an example of a workflow using paleobuddy to test inference methods.

As before, start with a simulation

```{r}
# we set a seed so the results are reproducible
set.seed(5)

# set the necessary parameters
# initial number of species
n0 <- 1

# speciation rate - it can be any function of time!
lambda <- function(t) {
  0.1 + 0.001*t
}

# extinction rate - also can be any function of time
mu <- function(t) {
  0.03 * exp(-0.01*t)
}

# maximum simulation time - species that die after this are considered extant
tMax <- 50

# run the simulation
sim <- bd.sim(n0, lambda, mu, tMax = tMax)

# check the resulting clade out
ape::plot.phylo(make.phylo(sim), show.tip.label = FALSE)
ape::axisPhylo()
```

A lot of species! We can then create a fossil record from this group.

```{r}
# again set a seed
set.seed(1)

# set the sampling rate
# using a simple case - there will be on average T occurrences,
# per species, where T is the species duration
rho <- 1

# bins - used to represent the uncertainty in fossil occurrence times
bins <- seq(tMax, 0, -1)
# this is a simple 1my bin vector, but one could use the GSA timescale
# or something random, etc

# run the sampling simulation only for the first 10 species for brevity's sake
# returnAll = TRUE makes it so the occurrences are returned as binned as well
# (e.g. an occurrence at time 42.34 is returned as between 42 and 41)
fossils <- suppressMessages(sample.clade(sim = sim, rho = rho, 
                                      tMax = tMax, S = 1:10, 
                                      bins = bins, returnAll = TRUE))
# suppressing messages - the message is to inform the user how
# many speciesleft no fossils. In this case, it was 0

# take a look at how the output is organized
head(fossils)
```

The output of `sample.clade` is a data frame organized as follows

* `Species` the species name, usually `t` followed by the number in the order it is organized on `sim`.
* `Extant` whether the species is extant or not.
* `SampT` the true occurrence time of a fossil occurrence. Returned if `returnTrue` and/or `returnAll` are set to `TRUE`.
* `MinT` the lower bound of the geologic range the fossil is found. The range vectors is an input, `bins`, used to simulate the granularity of the fossil record.
* `MaxT` the upper bound of the geologic range the fossil is found. The range columns are provided if `returnTRUE` is set to `FALSE` or if `returnAll` is set to `TRUE`.

We can visualize the fossil record of this group using `draw.sim` as well.

```{r}
# take only first 5 species with head
simHead <- head(sim, 10)

# draw longevities with fossil time points
suppressMessages(draw.sim(simHead, fossils = fossils))
```

We can also visualize the fossil ranges if we take away the `SampT` column, which is used by default if the data frame has it.

```{r}
# draw longevities with fossil time ranges
suppressMessages(draw.sim(simHead, fossils = fossils[, -3]))
```

This fossil record's data frame organization allows for easy evaluation of its components, and is close to ready for use in PyRate, one of the most widely used birth-death rates estimators in the field currently. To make it ready we must manipulate the `Extant` column, left like this for clarity. In PyRate, that column must be `status`, with extant species marked `extant` and extinct species marked `extinct`.

```{r}
# make a copy
pFossils <- fossils[, -3]

# change the extant column
pFossils["Extant"][pFossils["Extant"] == FALSE] = "extant"
pFossils["Extant"][pFossils["Extant"] == TRUE] = "extinct"

# change column names
colnames(pFossils) <- c("Species", "Status", "min_age", "max_age")

# check it out
head(pFossils)
```

This data frame could be directly dropped in PyRate for rate estimations. 

We can also use `RMark` or other simpler methods to check estimations. Let us write a quick function applying Foote's Per Capita method.

```{r}
per.capita <- function(faBins, laBins, bins) {
  # create vectors to hold species that were born before and die in interval i,
  # species who were born in i and die later,
  # and species who were born before and die later
  NbL <- NFt <- Nbt <- rep(0, length(bins))
  
  # for each interval
  for (i in 1:length(bins)) {
    # number of species that were already around before i and are not seen again
    NbL[i] <- sum(faBins > bins[i] & laBins == bins[i])
    
    # number of species that were first seen in i and are seen later
    NFt[i] <- sum(faBins == bins[i] & laBins < bins[i])
    
    # number of species that were first seen before i and are seen after i
    Nbt[i] <- sum(faBins > bins[i] & laBins < bins[i])
  }
  
  # calculate the total rates
  p <- log((NFt + Nbt) / Nbt)
  q <- log((NbL + Nbt) / Nbt)
  return(list(p = p, q = q))
}
```

Let us first run simulations with more species

```{r}
set.seed(1)

# constant lambda and mu
lambda = 0.15
mu = 0.05

# nFinal means there will be at least 100 species in the simulation
sim <- bd.sim(n0, lambda, mu, tMax, nFinal = c(100, Inf))

# sample--note the higher rho, since this method relies strongly on perfect sampling
fossils <- suppressMessages(sample.clade(sim, rho = 10, tMax, bins = bins, returnTrue = FALSE))
```

To apply this function, we need to manipulate `fossils` a little bit

```{r}
# get the species names
ids <- unique(fossils$Species)

# get the first appearance bins - the first time in bins where the fossil was seen (lower bound)
faBins <- unlist(lapply(ids, function(i) max(fossils$MaxT[fossils$Species == i])))

# get the last appearance bins - last time in bins where the fossil was seen (upper bound)
laBins <- unlist(lapply(ids, function(i) min(fossils$MinT[fossils$Species == i])))

# get the estimates
pc <- per.capita(faBins, laBins, bins)

# get a mean for each rate
print(paste0("Speciation estimate: ", mean(pc$p[is.finite(pc$p)], na.rm = TRUE)))
print(paste0("Extinction estimate: ", mean(pc$q[is.finite(pc$q)], na.rm = TRUE)))

# and plots
plot(x = rev(bins), y = pc$p, type = "l", xlab = "Time (Mya)", ylab = "Speciation rate")
abline(h = lambda)

plot(x = rev(bins), y = pc$q, type = "l", xlab = "Time (Mya)", ylab = "Extinction rate")
abline(h = mu)
```

This method assumes perfect sampling, so estimaates are not perfect. But one can clearly see the trend towards the correct values.

Other examples of methods using only the fossil record are Capture-Mark-Recapture (Liow & Nichols 2010) and the Alroy 3-timer (Alroy 2014). One could also use `paleobuddy` to test methods that integrate fossil records and phylogenies, such as the Fossilized birth-death model (Stadler et al 2010, Heath et al 2014). While a thorough test of these methods is not in the scope of this vignette, we believe to have shown the workflow required for such tests.

## paleobuddy's flexibility

While showing use cases of the package is useful, we should also spend some time detailing the possible customization options in `bd.sim` and `sample.clade`.

We have already shown speciation and extinction rates can be any function of time, or constant numbers. Let us check some other possibilities out.

```{r}
# set a seed
set.seed(2)

# parameters to set things up
n0 <- 1
tMax <- 20

# speciation can be dependent on a time-series variable as well as time
lambda <- function(t, env) {
  0.01 * t + 0.01*exp(0.01*env)
}

# let us use the package's temperature data
data(temp)
# this could instead  be data(co2), the other environmental
# data frame supplied by paleobuddy

# we can make extinction be age-dependent by creating a shape parameter
mu <- 10
mShape <- 0.5
# this will make it so species durations are distributed
# as a Weibull with scale 10 and shape 2

# run the simulation
# we pass the shape and environmental parameters
sim <- suppressMessages(bd.sim(n0, lambda, mu, tMax = tMax, 
                               mShape = mShape, envL = temp))
# note that lShape and envM also exist
# the defaults for all of these customization options is NULL

# check out the phylogeny
ape::plot.phylo(make.phylo(sim), show.tip.label = FALSE)
ape::axisPhylo()
```

So as to avoid repetition, we will not run simulations for the next possibilities, but we will list them

```{r}
# speciation may be a step function, presented
# as a rates vector and a shift times vector
lList <- c(0.1, 0.2, 0.05)
lShifts <- c(0, 10, 15)
# lShifts could also be c(tMax, tMax - 10, tMax - 15) for identical results

# make.rate is the function bd.sim calls to create a 
# ratem but we will use it here to see how it looks
lambda <- make.rate(lList, tMax = 20, rateShifts = lShifts)
t <- seq(0, 20, 0.1)
plot(t, rev(lambda(t)), type = 'l', main = "Step function speciation rate",
     xlab = "Time (Mya)", ylab = "Speciation rate", xlim = c(20, 0))
```
```{r}
# it is not possible to combine the lambda(t, env) and lList methods listed above
# but we can create a step function dependent on environmental data with ifelse
mu_t <- function(t, env) {
  ifelse(t < 10, env,
         ifelse(t < 15, env * 2, env / 2))
}

# pass it to make.rate with environmental data
mu <- make.rate(mu_t, tMax = tMax, envRate = temp)
plot(t, rev(mu(t)), type = 'l', main = "Environmental step function extinction rate",
     xlab = "Time (Mya)", ylab = "Rate", xlim = c(20, 0))
```

Note all these customization options could serve for a Weibull scale as well, if the user wishes to have age-dependent speciation and/or extinction. The possibility of age-dependency with time-varying rates is, to our knowledge, exclusive to `paleobuddy`.

```{r eval=FALSE}
# set seed again
set.seed(1)

# age-dependent speciation with a step function
lList <- c(10, 5, 12)
lShifts <- c(0, 10, 15)
lShape <- 2

# age-dependent extinction with a step function of an environmental variable
q <- function(t, env) {
  ifelse(t < 10, 2*env,
         ifelse(t < 15, 1.4*env, env / 2))
}

# note shape can be time-dependent as well, though
# we advise for variation not to be too abrupt due
# to computational issues
mShape <- function(t) {
  return(1.2 + 0.025*t)
}

# run the simulation
sim <- suppressMessages(bd.sim(n0, lList, q, tMax = tMax, lShape = lShape, 
                               mShape = mShape, 
                               envM = temp, lShifts = lShifts))

# check out the phylogeny
ape::plot.phylo(make.phylo(sim), show.tip.label = FALSE)
```

We set this to not run because it takes a long time - both because of the high speciation and because the method of creating a step function rate with `shifts` and `lists` takes a long time to integrate. It is however a good illustration of the power of `paleobuddy`s flexibility. 

None of the scenarios we have listed here are specific to speciation or extinction - one can mix and match any combination listed here for either rate. 

In the case of sampling rates, it is almost the case that any scenario available to speciation/extinction is available to sampling. We do not allow for a `shape` parameter since the Weibull distribution is not frequently used in the literature for age-dependent sampling. Instead, we allow for the user to supply a probability distribution describing how occurrences are distributed along a species age, as follows

```{r}
# as an example, we will use a PERT distribution, 
# a hat-shaped distribution used in PyRate

# preservation function
dPERT <- function(t, s, e, sp, a = 3, b = 3, log = FALSE) {

  # check if it is a valid PERT
  if (e >= s) {
    message("There is no PERT with e >= s")
    return(rep(NaN, times = length(t)))
  }

  # find the valid and invalid times
  id1 <- which(t <= e | t >= s)
  id2 <- which(!(t <= e | t >= s))
  t <- t[id2]

  # initialize result vector
  res <- vector()

  # if user wants a log function
  if (log) {
    # invalid times get -Inf
    res[id1] <- -Inf

    # valid times calculated with log
    res[id2] <- log(((s - t) ^ 2)*((-e + t) ^ 2)/((s - e) ^ 5*beta(a,b)))
  }
  
  # otherwise
  else{
    res[id1] <- 0

    res[id2] <- ((s - t) ^ 2)*((-e + t) ^ 2)/((s - e) ^ 5*beta(a,b))
  }

  return(res)
}

# set seed
set.seed(1)

# generate a quick simulation
sim <- bd.sim(n0, lambda = 0.1, mu = 0.05, tMax = 20)

# sample for the first 10 species
fossils <- suppressMessages(sample.clade(sim = sim, rho = 3, 
                                      tMax = 20, adFun = dPERT))
# here we return true times of fossil occurrences

# draw longevities with fossil occurrences
draw.sim(sim, fossils = fossils)
```

Note how occurrences cluster in the middle of a species' duration, as expected since the PERT is a hat-shaped distribution.

For more details and examples of age-dependent models, check out `?sample.clade`. Even if an `adFun` argument is supplied for age dependency, the average sampling rate `rho` may have the same flexibility as speciation and extinction rates in the birth-death functions. 

We provide below a final example containing a lot of complexity on both birth-death and fossil sampling rates. While set not to run due to its long runtime, users are encouraged to run it on their own time, and use it as a basis for the exploration of other sources of variation for rates (e.g. try using the PERT to make fossil sampling age-dependent, as above). It is completely self-contained, as long as paleobuddy is attached.

```{r eval=FALSE}
## setup
# set seed
set.seed(6)

# initial number of species
n0 <- 1

# maximum simulation time
tMax <- 30

# lambda varies with absolute time (exponential decay)
lambda <- function(t) {
  1 + exp(-t/2)
}

# mu will vary with species age and environmental variables
data(co2)
data(temp)

# its scale will depend on CO2, except for a mass extinction event
scale <- function(t, env) {
  ifelse(t < 10, env/20,
         ifelse(t < 11, 1/1.9, env/30))
  # ME between 10 and 11my from simulation start
  # step function-like dependence on CO2 otherwise
}
mu <- make.rate(scale, tMax = tMax, envRate = co2)

# its shape will vary with time and temperature
shape <- function(t, env) {
  0.5 + 2.5 * log(t + 1) / env
}
mu_shape <- make.rate(shape, tMax = tMax, envRate = temp)

# finally, fossil sampling will vary linearly with absolute time
rho <- function(t) {
  0.1 + 0.1 * t
}

## plot diversification rates to visualize
# time vector
time <- seq(0, tMax, by = 0.01)

# plotting lambda, mu scale, and mu shape
plot(time, rev(lambda(time)), type = "l", xlim = c(tMax, 0), ylim = c(0, 2),
     xlab = "Time (Mya)", ylab = "Rate value", 
     main = "Speciation rate, extinction scale and shape through time",
     col = "orange")
lines(time, rev(1/mu(time)), col = "red")
lines(time, rev(mu_shape(time)), col = "blue")
legend(x = 15, y = 0.7, lty = 1, col = c("orange", "red", "blue"),
       legend = c("lambda", "1/(mu scale)", "mu shape"))
# note we plotted the inverse of scale, since Weibull scales are
# equivalent to exponential rates when shape is equal to 1

## simulate
# birth-death simulation
sim <- suppressMessages(bd.sim(n0, lambda, mu, tMax, mShape = mu_shape))

# fossil record simulation - returning true sampling times
fossils <- suppressMessages(sample.clade(sim, rho, tMax, returnTrue = TRUE))

# draw BD simulation with fossils
draw.sim(sim, fossils = fossils)

# draw phylogenetic tree
ape::plot.phylo(make.phylo(sim))
```

## Conclusion

For completion, we present a final list of the functions of the package

* `bd.sim` birth-death simulator. Rates can be constant, time dependent, time and environmentally dependent, or a list of numbers. User can supply a shape parameter to make rate a Weibull scale.
* `make.phylo` creates a `phylo` object from the `ape` package.
* `sample.clade` simulates fossil samples from a set of species, usually outputted from `bd.sim`. Sampling rate can be anything a speciation or extinction rate can be, without the option of a shape parameter. The user may instead supply a distribution of occurrences across a species age, so as to simulate age-dependency in sampling with more flexibility.
* `draw.sim` function to draw longevities of the species in a `sim` object and, optionally, fossil occurrences of these species.
* `rexp.var` exponential and Weibull waiting time drawing with variable rates. Used by `bd.sim` and `sample.clade`.
* `find.lineages` separates a group returned by the `bd.sim` function, or similar, by the species that started the simulation as mothers of each group. Allows for an optional argument, a list of numbers, and if supplied returns the groups generated by those species instead. A quick example:

```{r}
# set a seed 
set.seed(1)

# simple simulation, starting with more than one species
sim <- bd.sim(n0 = 2, lambda = 0.1, mu = 0, tMax = 20, nFinal = c(20, Inf))

# separate the lineages
clades <- find.lineages(sim)

# plot each phylogeny

# clade 1
ape::plot.phylo(make.phylo(clades$clade_1$sim), show.tip.label = FALSE)

# clade 2
ape::plot.phylo(make.phylo(clades$clade_2$sim), show.tip.label = FALSE)
```

* `phylo.to.sim` creates a `sim` object from a phylogeny following the APE `phylo` class format (Paradis et al 2004). Can be used to integrate paleobuddy with other packages that output phylogenies. Note that the user must supply some optional arguments to allow for this; most importantly the information of each species' mother, since this is ambiguous from a bifurcating phylogeny.
* `make.rate` creates a function of time from the customization options presented above - vector of numbers and vector of shift times, function of time and environment and an environment matrix, etc. Used by `bd.sim` and `sample.clade` to detect when the given rates are constant, and create the rates to pass to the helper functions.
* `var.rate.div` calculates the expected diversity from a variable rate birth-death process and a time period.
* `binner` given occurrence times and time bins, returns the number of occurrences in each bin.
* Finally, there are a number of generics and methods for the `sim` class, like `print.sim` and `plot.sim`. See `?sim` for a detailed list.

While the package is robust and flexible, it is useful to spend a little time discussing the shortcomings and planned features.

* Currently, the possibility of using a time-varying Weibull shape is implemented, but failing certain tests when shape varies too much or is too low (see `?bd.sim` for a summary of issues). This seems to be due to the nature of R numerical integration, so we need to find workarounds for testing this feature more accurately.
* The method of generating step functions from rate and shift times vectors is currently very slow (see `?make.rate`), and we plan on looking for more efficient alternatives. Meanwhile, the user might prefer the creation of step functions using `ifelse` when prioritizing efficiency.
* A couple of diversification scenarios are very well established in the literature and therefore are planned for future versions of the package: trait-dependent diversification, diversity-dependent rates, anagenesis/bifurcating speciation, etc.

While there are a number of areas where `paleobuddy` can improve, it is clear that the package presents unprecedented flexibility on diversification, fossil record, and phylogeny generation. It will therefore be an impactful tool in the exploration of complex evolutionary scenarios.
